Lawvere-Tierney Sheafification\\ in Homotopy Type Theory

### 2015-01-14 paper/sheaf_lics.tex
work on introduction

The sheaf construction, or sheafification~\cite{maclanemoerdijk}, is a
very powerful tool to construct new topoi from already existing ones,
allowing logical principles---that can not be proven to be true or false
in the old topos---to be valided (or invalided) in the new topos.
%
The most common use of sheafification is to construct a boolean
topos, thus validating the classical principle of the law of excluded
middle (EM), from an intuitionistic topos. 
%
An other famous (revisited) application has been developed by Paul
Cohen~\cite{cohen1966} to prove that the continuum hypothesis is
independent of the usual axioms of Zermelo-Fr\ae nkel set theory, even
in presence of the axiom of choice (AC).

Even if similar questions have been considered around the Curry-Howard
isomorphism, there is no canonical way to extend a programming language
such as type theory while keeping consistency automatically. 
%
For instance, much effort have been done to provide a computational
content to the law of excluded middle in order to define a constructive
version of classical logic. This has lead to various calculi, with
most notably the $\lambda \mu$-calculus of
Parigot~\cite{parigot1993classical}, but this has not allow to define
a new version of type theory with classical principles.
%
Other works have tried to extend continuation-passing-style (CPS)
transformation to type theory, but they have been faced with the
difficulty that the CPS transformation is incompatible with (full) dependent
sums~\cite{barthe2002cps}, which puts emphasis on the tedious link
between the axiom of choice and the law of excluded middle in type theory.

Nevertheless the axiom of choice Coquand, Herbelin, Krivine.
%
% Some works have tried to provide a realization of the axiom of choice 
% compatible with classical logic

\begin{itemize}
\item 
  links between HoTT and higher topos theory 
\item 
  modalities in HoTT
\item 
  Giraud's axiom
\end{itemize}

>>>>

Sheafification~\cite{maclanemoerdijk} is a very powerful geometric
construction that has been initially stated in topology and has
quickly been lifted to mathematical logic.
%
In the field of topos theory, it provides a way to construct new topoi
from already existing ones, allowing logical principles---that can not
be proven to be true or false in the old topos---to be valid (or
invalid) in the new topos.
%
The most common use of sheafification is the construction a boolean
topos, thus validating the classical principle of the law of excluded
middle (EM), from an intuitionistic topos using the dense topology. 
%
An other famous application has been developed by
Cohen~\cite{cohen1966} to prove that the continuum hypothesis is
independent of the usual axioms of Zermelo-Fr\ae nkel set theory, even
in presence of the axiom of choice (AC). The initial work of Cohen
uses forcing but can be rephrased in terms of
sheafification~\cite{maclanemoerdijk}.

Even if similar questions have been considered around the Curry-Howard
isomorphism, there is no canonical way to extend a programming
language such as type theory with new logical or computational
principle while keeping consistency automatically.
%
For instance, much effort have been done to provide a computational
content to the law of excluded middle in order to define a
constructive version of classical logic. This has lead to various
calculi, with most notably the $\lambda \mu$-calculus of
Parigot~\cite{parigot1993classical}, but this line of work has not
appeared to be fruitful to define a new version of type theory with
classical principles.
%
Other works have tried to extend continuation-passing-style (CPS)
transformation to type theory, but they have been faced with the
difficulty that the CPS transformation is incompatible with (full) dependent
sums~\cite{barthe2002cps}, which puts emphasis on the tedious link
between the axiom of choice and the law of excluded middle in type theory.

Nevertheless the axiom of choice has shown to be realizable by
computational meaning in a classical setting by techniques turning
around the notion of (modified) bar induction
\cite{berardi1998computational}, Krivine's
realizability~\cite{krivine2003dependent} and even more recently with
restriction on elimination of dependent sums and lazy
evaluation~\cite{herbelin2012constructive}.
%

Unfortunately, all those works do not share the generality and
simplicity of sheafification in topos theory.  
%
However, type theory is known to be quite closed to topos theory so
one could wonder why similar techniques have not been developed in the
field of type theory. 
%
The answer to this question has been given recently by the advent of
homotopy type theory~\cite{hottbook}, which is an extension of
Martin-Löf type theory with principles inspired by (higher) category
theory and homotopy theory, such as higher inductive
types~\cite{lumsdaine2011higher,lumsdaine2013higher} and
Voevodsky's univalence principle~\cite{kapulkin2012simplicial}.

This new point of view on type theory has revealed the
homotopy structure of types where for instance (homotopical)
propositions are just types with an irrelevant equality and
(homotopical) sets are types with a propositional equality. 
%
When restricted to such propositions and sets, type theory corresponds
quite closely to topos theory but the mismatch starts when considering
higher homotopical types. 
%
Fortunately, an higher version of topos theory has been developed
recently by mathematicians, synthesized in the monograph of Lurie on
higher topos theory~\cite{lurie}. 
%
Lurie has developed their all the tools that have been defined in
topos theory, but in an higher setting. In particular, the theory of
sheaves has been lifted to higher topos theory.
%
As the notion of higher topoi appears to correspond very closely
to homotopy type theory, this provides a new hope that tackling the problem of
extending the power of type theory using sheafification is actually possible.

Nevertheless, the adaptation of the sheafification in higher topos
theory to homotopy type theory is not straightforward because the
construction in higher topoi is restricted to the initial
Grothendieck which is still very topological, and not very amenable to
formalization in type theory. It seems more promising to use a
generalized sheafification, called Lawvere-Tierney
sheafification~\cite{maclanemoerdijk}, but this construction has not
been considered yet in the setting of higher topos theory.

\subsection{Lawvere-Tierney Topology}

\subsection{Overview of the Result}

The main contribution of this article is presents a computer check

\begin{itemize}
\item 
  modalities in HoTT
\item 
  Giraud's axiom
\end{itemize}
